---
title: "PCA_signalanalysis_mantel"
output: html_document
---
# set working directory
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# clean list
```{r}
rm(list=ls())
```

# load data and packages
```{r}
require(plyr)
require(ggplot2)
require(reshape2)
require(ggfortify)
require(factoextra)
require(vegan)
library(hexbin)
library(tidyr)

#Load EOD measurements made in matlab
eoddata<-read.csv("kingsEODdatafromgenotyping_2016.csv",header=TRUE,sep=",")
```

#Data formatting and PCA analysis
```{r}
#script never measured these variables, so drop them
drops<-c("tS0","vs0","sS0")
eoddata<-eoddata[,!names(eoddata) %in% drops]
eoddata<-na.omit(eoddata)

#zscore the data (as per Original MATLAB script)
eoddata_scaled<-scale(eoddata[,10:29],center=TRUE,scale=TRUE)

#do the PCA analysis
pcadata<-princomp(eoddata_scaled)

#Get factor loadings for Table S3
res.var2 <- get_pca_var(pcadata)
factor_loadings2=res.var2$contrib
factor_loadings2
write.csv(factor_loadings2, 'factor_loadings_tableS3.csv')

pcadata_df<-data.frame(eoddata$fn,eoddata$Valid_Loc_Short,pcadata$scores)
colnames(pcadata_df)<-c("fn", "loc","c1","c2","c3","c4","c5","c6","c7","c8","c9","c10","c11","c12","c13","c14","c15","c16","c17","c18","c19","c20")

#Plot the PCA
PCAloc=autoplot(pcadata, scale=0, data=pcadata_df, colour="loc", size=2, frame=TRUE, frame.colour="loc", fill=NULL) +theme_classic()
PCAloc

#change colors
PCAloc=PCAloc + scale_color_manual(values=c("#b30000","#feb24c","#8c96c6","#41b6c4", "#addd8e","#238443","#ec7014","#c51b8a","#e31a1c","#fa9fb5","#081d58"))
```

#calculate centroids of population polygons in the PCA and make Euclidean EOD distance matrix (PCA-derived signal distances) for Gabon and for the subset of Southern populations
```{r}
pcadata_df<-data.frame(eoddata$fn,eoddata$Valid_Loc_Short,pcadata$scores)
colnames(pcadata_df)<-c("fn", "loc","c1","c2","c3","c4","c5","c6","c7","c8","c9","c10","c11","c12","c13","c14","c15","c16","c17","c18","c19","c20")

meanphenotype<-aggregate(pcadata_df[,3:4],list(pcadata_df$loc),mean)
rownames(meanphenotype)=meanphenotype$Group.1
loc=rownames(meanphenotype)
loc=as.factor(loc)

#Euclidean PCA-derived signal distances for the whole country
phenodist<-vegdist(meanphenotype[,2:3], method="euclidean")
write.csv(as.matrix(phenodist), "phenodist_PCA.csv")

#Euclidean PCA-derived signal distances for the Subset of Southern populations
meanphenotypesouth<-meanphenotype[-c(2,9,10),] # remove 
phenodist_south<-vegdist(meanphenotypesouth[,2:3], method="euclidean")
write.csv(as.matrix(phenodist_south), "phenodist_PCA_south.csv")
```

#add population centroids to the PCA plot
```{r}
PCAloc_centroids=PCAloc + geom_point(data=meanphenotype, aes(x=c1, y=c2), size=2)
PCAloc_centroids

ggsave("PCA_loc_centroids.eps", width=30, height=20, units="cm")
```

#Perform PERMANOVA analysis on PCA-derived signal distances
```{r}
pheno_ind_scores<-pcadata_df[,3:4]
rownames(pheno_ind_scores)=pcadata_df$fn
loc<-factor(pcadata_df$loc)
PCA_permanova=adonis(pheno_ind_scores ~ loc, permutations=1000, method='eu')
PCA_permanova
```

#Perform PERMANOVA analysis on MDS-derived signal distances
```{r}
#load the MDS-derived EOD coordinates
mdscoordinates=read.csv("mds_eodcoordinates.csv", header=TRUE)
rownames(mdscoordinates)=1:nrow(mdscoordinates)
mdscoordinates=mdscoordinates[order(mdscoordinates$pop),]
mdscoordinates_final=mdscoordinates[,3:4]
loc=mdscoordinates[,2]
loc=as.factor(loc)
MDS_permanova=adonis(mdscoordinates_final~ loc, permutations=1000, method='eu')
MDS_permanova
```

#Mantel tests analyses on PCA-derived signal distances: whole Gabon
```{r}
#load the genetic, geographic, and regional cluster matrices
gen1<-read.csv("king_microsats_rousset.csv", header=TRUE)
row.names(gen1)<-gen1$X
gen1$X<-NULL
gendist<-as.dist(gen1)

geo<-read.csv("geo_dist.csv", header=TRUE, row.names=1, check.names=FALSE)
geo=as.matrix(geo)
geo.dist=as.dist(geo)

reg<-read.csv("regional_effects.csv", header=TRUE, row.names=1, check.names=FALSE)
reg=as.matrix(reg)
regdist=as.dist(reg)
regdist[is.na(regdist)] <- 0


#Genetic-geographic
mantel(gendist,geo.dist, permutations=1000)

#Signal-genetic
mantel(gendist,phenodist, permutations=1000)

#Signal-geographic
mantel(phenodist, geo.dist, permutations=1000)

#Partial correlation
#genetic - geographic (regional clusters as covariate)
mantel.partial(gendist,geo.dist, regdist, permutations=1000)

#signal - genetic         (regional clusters as covariate)
mantel.partial(phenodist, gendist,regdist, permutations=1000)

#signal - geographic   (regional clusters as covariate)
mantel.partial(phenodist, geo.dist, regdist, permutations=1000)
```

#Mantel test analyses on MDS-derived signal distances: Whole Gabon
```{r}
#load the MDS-derived centroids of population polygons and make Euclidean EOD distance matrix (MDS-derived signal distances) for Gabon an
mds_centroids=read.csv("mds_centroids.csv", header=TRUE)
mds_phenodist=vegdist(mds_centroids[,-1], method="euclidean")
rownames(mds_centroids)=mds_centroids$X

#Genetic-geographic
mantel(gendist,geo.dist, permutations=1000)

#Signal-genetic
mantel(gendist,mds_phenodist, permutations=1000)

#Signal-geographic
mantel(mds_phenodist, geo.dist, permutations=1000)

#Partial correlation
#genetic - geographic (regional clusters as covariate)
mantel.partial(gendist,geo.dist, regdist, permutations=1000)

#signal - genetic         (regional clusters as covariate)
mantel.partial(mds_phenodist, gendist,regdist, permutations=1000)

#signal - geographic   (regional clusters as covariate)
mantel.partial(mds_phenodist, geo.dist, regdist, permutations=1000)

```

#Mantel test analyses on PCA-derived signal distances: South
```{r}
#load the genetic, geographic, and Bongolo Falls matrix
gen1s<-read.csv("king_microsats_rousset_south.csv", header=TRUE)
row.names(gen1s)<-gen1s$X
gen1s$X<-NULL
gendists<-as.dist(gen1s)

geos<-read.csv("geo_dist_south.csv", header=TRUE, row.names=1, check.names=FALSE)
geos=as.matrix(geos)
geo.dists=as.dist(geos)

bongolo<-read.csv("pop_bongolo_matrix.csv", header=TRUE, row.names=1, check.names=FALSE)
bongolo=as.matrix(bongolo)
bongolo=as.dist(bongolo)

#Genetic-geographic
mantel(gendists,geo.dists, permutations=1000)

#Genetic-Bongolo
mantel(gendists, bongolo, permutations=1000)

#Signal-genetic
mantel(gendists,phenodist_s, permutations=1000)

#Signal-geographic
mantel(phenodist_s, geo.dists, permutations=1000)

#signal-Bongolo
mantel(phenodist_s, bongolo, permutations=1000)
```

#Mantel test analyses on MDS-derived signal distances: South
```{r}
# make Euclidean EOD distance matrix (MDS-derived signal distances) for Southern populations
mds_centroids_south<-mds_centroids[-c(2,9,10),]
mds_phenodists=vegdist(mds_centroids_south[,-1], method="euclidean")

#Genetic-geographic
mantel(gendists,geo.dists, permutations=1000)

#Genetic-Bongolo
mantel(gendists, bongolo, permutations=1000)

#Signal-genetic
mantel(gendists,mds_phenodists, permutations=1000)

#Signal-geographic
mantel(mds_phenodists, geo.dists, permutations=1000)

#signal-Bongolo
mantel(mds_phenodists, bongolo, permutations=1000)
```
